Causal Discovery on Precarity and Depression

### 2025-01-16 draft_v1/draft_v1.qmd
update_pc_sym

```{r}
#| layout-nrow: 2
#| label: fig-pc_sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
#| fig-subcap:
#|   - "Using both GaussianCI and RCoT"
#|   - "Using only RCoT"
#| fig-align: center
#| out-width: 70%

knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png")
knitr::include_graphics("img/pc_sym_onlyRCoT.png")
```

>>>>

<!-- ```{r} -->
<!-- #| layout-nrow: 2 -->
<!-- #| label: fig-pc_sym -->
<!-- #| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC." -->
<!-- #| fig-subcap: -->
<!-- #|   - "Using both GaussianCI and RCoT" -->
<!-- #|   - "Using only RCoT" -->
<!-- #| fig-align: center -->
<!-- #| out-width: 70% -->

<!-- knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png") -->
<!-- knitr::include_graphics("img/pc_sym_onlyRCoT.png") -->
<!-- ``` -->

```{r}
#| label: fig-pc_sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
#| fig-align: center
#| out-width: 70%

knitr::include_graphics("img/PC_symptom.png")
```

### 2025-01-16 draft_v1/draft_v1.tex
update_pc_sym

\date{2025-01-12}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_sym_GaussianCIRCoT.png}

@@

\subcaption{\label{fig-pc_sym-1}Using both GaussianCI and RCoT}

\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_sym_onlyRCoT.png}

\subcaption{\label{fig-pc_sym-2}Using only RCoT}

>>>>

\date{2025-01-16}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_symptom.png}

@@
